---
title: "Nuclear Weapon Tests Data Visualization"
author: "Mariaana Talus"
date: "`r format(Sys.time(), '%d.%m.%Y')`" #today 
output: html_document 
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.html'))})
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Nuclear Weapon Tests - Data Visualization

The goal is to present nuclear weapon test data on a map. But first, an extensive data cleaning was needed, see document NuclearWeaponTests_Cleaning.Rmd. Together with the map, different plots are presented to present differences and similarities between tests by different testing parties. 

## Import libraries and set constants 

Import libraries
```{r import libraries, message=FALSE}
#install.packages("plotly")  # needed only if plotly has not been installed already
library(plotly)
library(ggplot2)
library(dplyr)
```

Set some constants for the number of rows shown in tables.
```{r set constants}
ROWS = 30 # how many data rows are displayed 
```

## Reading in data

Reading in data. Data is downloaded from http://nuclearweaponarchive.org/Library/Catalog, and it has been saved as nuclearweapondata.txt. Updated version can be found from https://www.batguano.com/nuclear/nuccatalog.html. 

```{r read in data}
df_data <- read.csv("nuclearweapondata_cleaned.csv")
```


The following data is presented in the file: 

* Index
* Date
* Test Party
* Test Party Description
* Site
* Site Description
* Test Type
* Test Type Description
* Seismic Body
* Yield
* Latitude
* Longitude
* Purpose
* Purpose Description
* Device
* Device Description
* Name

A glipse of the data to confirm everything looks normal: 

```{r preview data}
head(df_data, 10) 
```

## Modify time for plots 

Change date to POSIXct format and add a year column
```{r data date}
df_data$Date <- as.POSIXct(df_data$Date)
df_data$Year <- format(df_data$Date,"%Y")
head(df_data$Year, 10) 
```

# Create Plots

## Test Locations on Map 

Create text to show when hovering on map. 

```{r map hovertext}
df_data$hover <- with(df_data, paste("Test Site:", Site_Descr, "<br>", "Date:", Date, '<br>', 
                                     "Test party:", TestParty_Descr, "<br>", "Name:", Name, '<br>', 
                                     "Device:", Device_Descr, "<br>", "Yield:", Yield, "<br>",
                                     "Purpose:", Purpose_Descr, "<br>", "Type:", TestType_Descr
                                      ))
```

Print map: 

```{r plot map}
g <- list( # define map elements  
  showland = TRUE,
  landcolor = "White",
  showocean = TRUE,
  oceancolor = "LightBlue",
  showlakes = TRUE,
  lakecolor = "LightBlue",
  showrivers = TRUE,
  rivercolor = "LightBlue",
  showcountries = TRUE,
  countrycolor = "DarkGray",
  resolution = 50,
  projection = list( type = 'natural earth')
)
# define data shown on map 
fig <- plot_geo(df_data, lat = ~Lat, lon = ~Lng)
fig <- fig %>% add_markers(text = df_data$hover, 
                           color = df_data$TestParty_Descr,
                           size = I(25))
# define layout 
fig <- fig %>% layout(
    title = 'Nuclear tests', 
    geo = g
  )
fig
```

Note that most locations on the map are approximate.  

## Tests by Testing Party

Get frequencies for the plot;  Get number of tests by testing party 

```{r count total events}
totCounts <- data.frame(df_data %>% group_by(TestParty_Descr) %>% summarise(NoEvents = n()))
totCounts <- na.omit(totCounts) # remove NaNs
totCounts
```

Create plot

```{r plot tot amounts}
fig <- plot_ly(totCounts, labels = ~TestParty_Descr, values = ~NoEvents, type = 'pie', 
               textinfo = "percent", hoverinfo = 'text', textposition = 'inside', 
               text = ~paste(TestParty_Descr, ': ', NoEvents))
fig <- fig %>% layout(title = 'Nuclear Tests by Testing Party')
fig
```

## Tests by Year and Testing Party

Get frequencies for the plot; Get tests by year and testing party. 

```{r count yearly events}
eventCounts <- data.frame(df_data %>% group_by(Year,TestParty_Descr) %>% summarise(NoEvents = n()))
head(eventCounts, 10)
```

Create plot

```{r plot dates}
p <- ggplot(eventCounts, aes(fill=TestParty_Descr, y=NoEvents, x=Year)) + 
     geom_bar(position="stack", stat="identity", width = 0.9) + 
     theme(legend.title = element_blank())
fig <- ggplotly(p)
fig <- fig %>% layout(title = 'Number of Nuclear Tests Yearly',
                      xaxis = list(tickangle = 90, title = ""),
                      yaxis = list(title = ""))
fig
```

## Test Types by Testing Party

Get tests by device type and testing party  
```{r count test types}
df_testTypes <- data.frame(df_data %>% group_by(TestType_Descr,TestParty_Descr) %>% summarise(noTestTypes = n()))
head(df_testTypes, 10)
```

Create plot 

```{r plot test types}
p <- ggplot(df_testTypes, aes(fill=TestParty_Descr, y=noTestTypes, x=TestType_Descr)) + 
     geom_bar(position="stack", stat="identity", width = 0.9) +
     theme(legend.title = element_blank())
fig <- ggplotly(p)
fig <- fig %>% layout(title = 'Number of Tests by Type and Testing Party',
                      xaxis = list(tickangle = 45,title = ""),  
                      yaxis = list(title = ""))
fig
```

## Test Purposes by Testing Party

Get tests by device type and testing party  
```{r count purpose}
df_purposes <- data.frame(df_data %>% group_by(Purpose_Descr,TestParty_Descr) %>% summarise(noPurposes = n()))
head(df_purposes, 10)
```

Create plot 

```{r plot purpose}
p <- ggplot(df_purposes, aes(fill=TestParty_Descr, y=noPurposes, x=Purpose_Descr)) + 
     geom_bar(position="stack", stat="identity", width = 0.9) +
     theme(legend.title = element_blank())
fig <- ggplotly(p)
fig <- fig %>% layout(title = 'Number of Tests by Purpose and Testing Party',
                      xaxis = list(tickangle = 45, title = ""),
                      yaxis = list(title = ""))
fig
```
